Implementation and assessment of tools for integrating forest roads into projections of the cumulative effects of disturbance on wildlife

### 2024-03-12 analysis/paper/paper.Rmd
Update fig 4
change to ILCP throughout

We focus on methods that link multiple target locations to an existing road network for strategic purposes by finding minimum-cost locations for new roadways that link landings within harvest cutblocks (i.e., log decks) to the existing road network, described by Dean [-@dean_finding_1997] as the multiple target access problem (MTAP). Specifically, we compare dynamic least cost path (DLCP) and minimum spanning tree approaches (MST) for solving the MTAP to an alternative iterative least cost path method developed by Hardy [-@hardy_forest_2019;-@hardy_landis-ii_2023] (Figure \@ref(fig:proj-meth-fig)).

The DLCP approach solves the MTAP by decomposing the problem into several smaller single target access problems (STAPs). For each landing target, the least cost path to the existing road network is solved using Dijkstra's shortest path algorithm [-@dijkstra_note_1959]. The algorithm is parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction. Each node is connected with up to eight neighbours, and the cost of construction along edges is weighted by the distance to neighbouring nodes and the average of the construction cost at the source and destination nodes (Table \@ref(tab:cost-tbl)). The simple underlying grid is less computationally costly than methods designed to support curvature or switchbacks required in operational road design [@stuckelberger_improved_2007]. The algorithm starts with the landing nearest to an existing road, and updates the existing road network and construction cost surface with each new road segment before moving on to the next nearest landing. The method was developed by Anderson and Nelson [-@anderson_projecting_2004] who noted that the projected network is sensitive to the ordering of targets and recommended a heuristic for connecting the targets and finding optimal branch locations to deal with this sensitivity.

@@

Hardy's method [-@hardy_forest_2019;-@hardy_landis-ii_2023] is similar to our DLCP method in that it breaks the MTAP down into many STAPs and finds the least cost path for each with the ordering of road construction determined by the distance from the existing road network. One difference is that Hardy uses the Dijkstra algorithm to solve a “single-source many-targets shortest path problem”, which is a special case of STAPs that finds the path between the harvested cell and one of multiple potential connection points on the road network. In contrast our algorithm first finds the closest point on the road network to the selected landing point and then uses the Dijkstra algorithm to solve least cost path between the two. However, since the cost of construction on existing roads is zero we expect that in practice these two approaches are similar and the algorithm will select the least cost location for the road to connect to the existing road network. It is possible that a sub-optimal road connection location is chosen by our algorithm when the existing road network is not fully connected and the closest point is on an existing road fragment that is not connected to the optimal part of the road network. Within cutblocks, Hardy's method iterates over each cell in the harvest area and builds a road if it is not within skidding distance of an existing road, while we create landing points to serve as target locations for road construction.

@@

The DLCP and MST methods connect landings within each cutblock to the existing road network, so an important first step is to select one or more landing locations in each block. We investigated the sensitivity of results to the pattern and density of landing locations by comparing results derived from the centroid of each cutblock to regular or randomly selected landing densities of 1 (low) and 10 (high) points per km^2^ (Figure \@ref(fig:proj-meth-fig)). We used default settings in the Hardy QGIS plugin except for skidding distance which was set to 200m. A skidding distance of 200 m corresponds to \~ 9 pts/km^2^. We compared two resolutions of the cost surface by aggregating the 1 ha resolution cost raster to a coarser 100 ha (1000 by 1000 m) resolution. In total, we compared 12 road projection method variants to a cutblocks only scenario with no projected roads (Table \@ref(tab:method-tbl)).

```{r method-tbl}
tbl2 <- "Projection method|Sample type|Sample density (pts/km^2^)|Resolutions
MST|Regular|1|1 ha and 100 ha
MST|Regular|10|1 ha and 100 ha
MST|Random|1|1 ha and 100 ha
MST|Random|10|1 ha and 100 ha
MST|Centroid|N/A|1 ha and 100 ha
DLCP|Regular|10|1 ha 
Hardy QGIS|N/A|~9|1 ha
Cutblocks only|N/A|N/A|1 ha

@@

To help make these road network projection methods more easily and widely useable, we implemented the DLCP and MST methods in the open source R package `roads` [@endicott_roads_2023]. The `roads` package is dependent on several other R packages that made this work possible, including, among others: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @noauthor_igraph_2022], `terra` for manipulating spatial raster data [@hijmans_terra_2023], and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022]. The package is available on CRAN (The Comprehensive R Archive Network). The source code for the package is available on GitHub (<https://github.com/LandSciTech/roads>), along with tutorials and documentation (<https://landscitech.github.io/roads/>).

@@

dlcpProj <- projectAll(tsbs = tsb,
                      paramTable = paramTable %>% 
                        filter(sampleType == "regular") %>% 
                        mutate(method = "dlcp"),
                      costSurface = tsaCost,
                      cutblocks = cutblocks,
                      existingRoads = exRoads,
                      fileLocation = here(data_path_drvd, "..", "for_fig"))

allProj <- bind_rows(mstProj, dlcpProj, .id = "method") %>%
  arrange(desc(sampleType), sampleDens)

@@

allProj <- bind_rows(
  allProj,
  tibble(sampleType = "", sampleDens = 0, method = "Hardy QGIS",
             output = list(here(data_path_drvd, "..", "for_fig", 
                                "HardyQGISTest.gpkg")))) %>% 
  mutate(mapTitle =  paste(ifelse(method == 1, "MST",
                                  ifelse(method == 2, "DLCP", method)),
                           sampleType,
                           ifelse(sampleDens == 1e-06 & 
                                    sampleType != "centroid", "low density", 
                                  ifelse(sampleDens == 1e-05, 
                                         "high density", ""))))

@@

```{r proj-meth-fig, fig.cap="Comparison of minimum spanning tree (MST), dynamic least cost paths (DLCP), and the Hardy QGIS plugin road network projection algorithms. Landing locations were selected as the centroid of the cutblock, or a random or regularly spaced sample of points inside each cutblock (Table 2). Target cutblocks are outlined in grey, existing roads are black, projected roads are yellow, observed roads are red. Cost (1 ha resolution) is indicated by grey shading. The large four sided cutblock in the top right was added to more clearly show algorithm behaviour within cutblocks."}
knitr::include_graphics(here("analysis/figures/projection_methods_figure.png"))
```

@@

```{r aspat-perf-fig, fig.height=4, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants within cutblocks, outside cutblocks, and overall (1 ha cost raster). Values greater than zero indicate overprojection and negative values indicate underprojection. Cutblocks not accessible via the observed road network were excluded from analysis. Performance varies among metrics (columns), projection method variants (colours), and within and outside cutblocks."}
# organize data for ggplot
doAspatPlot <- function(tab){
  meanTable_long <- tab %>% dplyr::select(-runTime, -order) %>% 
  pivot_longer(-c(sampleType, sampleDens, areaMean, method),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
                             sampleType),
         sampleDens = paste(method, sampleType, sampleDens) %>% 
           factor(levels = c("mst NA centroid",
                             "NA NA observed",
                             "mst random low sample density",
                             "mst regular low sample density",
                             "mst random high sample density",
                             "mst regular high sample density",
                             "dlcp regular high sample density",
                             "NA NA klementQGIS",
                             "NA NA NA"),
                  labels = c("MST centroid", 
                             "Observed",
                             "MST random low density",
                             "MST regular low density",
                             "MST random high density",
                             "MST regular high density",
                             "DLCP regular high density",
                             "Hardy QGIS",
                             "Cutblocks only")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road\ndensity", "Road\npresence",
                                    "Distance\nto road",
                                    "Forestry\ndisturbance\nfootprint",
                                    "Road\ndisturbance\nfootprint")))

@@

# save runTimes for later comparison
dlcp_time <- meanTableRealCuts %>% 
  filter(method == "dlcp", areaMean == "overall") %>% 
  pull(runTime) %>% round(2)

@@

knitr::include_graphics(here("analysis/figures/Figure4_diff_map_update.png"))

@@

```{r spat-perf-fig, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Aspatial variation of in performance among projection method variants (1 ha cost raster). F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted pixels that were observed and sensitivity is the proportion of observed pixels that were correctly predicted. Values closer to one indicate better performance. Cutblocks not accessible via the observed road network were excluded from analysis."}
doSpatPerf <- function(tab){
  matchData_sum <- tab %>% ungroup() %>%
    mutate(agreement = factor(agreement,
                              levels = c("False negative", "False positive",
                                         "Agree roaded","Agree roadless",
                                         "Pre-existing roads")),
           sampleDens = paste(method, sampleType, sampleDens) %>%
             str_replace("1e-06", "low density") %>%
             str_replace("1e-05", "high density") %>%
             str_replace("centroid low density", "centroid") %>%
             str_replace("klementQGIS NA", "klementQGIS") %>%
             factor(levels = c("mst centroid",
                               "mst random low density",
                               "mst regular low density",
                               "mst random high density",
                               "mst regular high density",
                               "dlcp regular high density",
                               "NA klementQGIS", 
                               "NA cutOnly NA"),
                    labels = c("MST centroid", 
                               "MST random low density",
                               "MST regular low density",
                               "MST random high density",
                               "MST regular high density",
                               "DLCP regular high density",
                               "Hardy QGIS", 
                               "Cutblocks only")),
           metric = factor(metric,
                           levels = c("roadPresence",
                                      "forestryDisturbance",
                                      "roadDisturbance"),
                           labels = c("Road presence",
                                      "Forestry disturbance\nfootprint",
                                      "Road disturbance\nfootprint"))) %>%
    group_by(metric, sampleDens, agreement) %>%
    summarise(count = sum(count), .groups = "drop_last") %>%
    mutate(perc = count/sum(count)*100)

@@

As expected, none of the projection methods yield accurate spatial projections of road presence (i.e. the exact locations of roads) (F-Measure: `r perf_tbl_rng["Road presence"]`). Neither do any of the methods provide better information about the exact location of the forestry disturbance footprint than the locations of cutblocks only (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`). All methods provide comparably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yield fewer false negatives and more false positives (Figure \@ref(fig:spat-perf-fig)). The DLCP projection method had comparable performance to the MST projection method with the same sampling strategy. The DLCP method was faster to compute however, with a run time of `r dlcp_time` minutes compared to `r mst_time` minutes for the MST method.

<!-- TO DO: Sarah figure out why `r dlcp_time` and `r mst_time` are coming out as NA, and how to fix that.-->

@@

Outside cutblocks, we saw very little variation in performance among projection method variants. All were somewhat better than nothing, but the improvement was small because in this landscape new landings were fairly close to existing roads. Outside cutblocks our DLCP algorithm is similar to Hardy's algorithm, so we expect little difference in behaviour where targets are far from existing roads. Within cutblocks, the density of landing points had a much larger impact on performance than the choice of DCLP or MST algorthm. The Hardy algorithm most accurately projected high road density within blocks, but examination of spatial patterns clarified that the projected tortuousity is quite different from observed switchbacks, and may be an artefact of searching for nearest destinations >200 m away on a 100 m grid. One could increase the density of landing points in the DCLP/MST projections to match the observed high density of roads within cutblocks, but if realistic switchbacks in mountainous terrain is a key goal then none of these algorithms are sufficient.

@@

meanTable <- meanTable %>% mutate(resolution = "100") %>% bind_rows(meanTable2) %>% 
  filter((method != "dlcp"| is.na(method)) & (sampleType != "cutOnly" | is.na(sampleType)))

matchData <- matchData %>% mutate(resolution = "100") %>% bind_rows(matchData2) %>% 
  filter(method != "dlcp", sampleType != "cutOnly")

@@

```{r figA1, fig.height=6, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods for the fine resolutiom (1 ha) and coarse resolution (100 ha) cost rasters. Metrics were summarised within cutblocks, outside cutblocks, and overall. Values greater than 0 indicate the projection overestimated the metric and negative values indicate an underestimation."}
# organize data for ggplot
meanTable_long <- meanTable %>% dplyr::select(-runTime, -order) %>% 
  pivot_longer(-c(sampleType, sampleDens, areaMean, resolution, method),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
                             sampleType),
         sampleDens = paste(method, sampleType, sampleDens) %>% 
           factor(levels = c("mst NA centroid",
                             "NA NA observed",
                             "mst random low sample density",
                             "mst regular low sample density",
                             "mst random high sample density",
                             "mst regular high sample density",
                             "dlcp regular high sample density",
                             "NA klementQGIS", 
                             "NA NA NA"),
                  labels = c("MST centroid", 
                             "Observed",
                             "MST random low density",
                             "MST regular low density",
                             "MST random high density",
                             "MST regular high density",
                             "DLCP regular high density",
                             "Hardy QGIS",
                             "Cutblocks only")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road\ndensity", "Road\npresence",
                                    "Distance\nto road",
                                    "Forestry\ndisturbance\nfootprint",
                                    "Road\ndisturbance\nfootprint")))

@@

matchData_sum <- matchData %>% ungroup() %>%
  mutate(agreement = factor(agreement,
                            levels = c("False negative", "False positive",
                                       "Agree roaded","Agree roadless",
                                       "Pre-existing roads")),
         sampleDens = paste(method, sampleType, sampleDens) %>%
           str_replace("1e-06", "low density") %>%
           str_replace("1e-05", "high density") %>%
           str_replace("centroid low density", "centroid") %>%
           str_replace("klementQGIS NA", "klementQGIS") %>%
           factor(levels = c("mst centroid",
                             "mst random low density",
                             "mst regular low density",
                             "mst random high density",
                             "mst regular high density",
                             "dlcp regular high density",
                             "NA klementQGIS"),
                  labels = c("MST centroid", 
                             "Random low density",
                             "Regular low density",
                             "Random high density",
                             "Regular high density",
                             "DLCP regular high density",
                             "Hardy QGIS")),
         metric = factor(metric,
                         levels = c("roadPresence",
                                    "forestryDisturbance",
                                    "roadDisturbance"),
                         labels = c("Road presence",
                                    "Forestry disturbance\nfootprint",
                                    "Road disturbance\nfootprint"))) %>%
  group_by(metric, sampleDens, agreement, resolution) %>%
  summarise(count = sum(count), .groups = "drop_last") %>%
  mutate(perc = count/sum(count)*100)

>>>>

We focus on methods that link multiple target locations to an existing road network for strategic purposes by finding minimum-cost locations for new roadways that link landings within harvest cutblocks (i.e., log decks) to the existing road network, described by Dean [-@dean_finding_1997] as the multiple target access problem (MTAP). Specifically, we compare iterative least cost path (ILCP) and minimum spanning tree approaches (MST) for solving the MTAP to an alternative iterative least cost path method developed by Hardy [-@hardy_forest_2019;-@hardy_landis-ii_2023] (Figure \@ref(fig:proj-meth-fig)).

The ILCP approach solves the MTAP by decomposing the problem into several smaller single target access problems (STAPs). For each landing target, the least cost path to the existing road network is solved using Dijkstra's shortest path algorithm [-@dijkstra_note_1959]. The algorithm is parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction. Each node is connected with up to eight neighbours, and the cost of construction along edges is weighted by the distance to neighbouring nodes and the average of the construction cost at the source and destination nodes (Table \@ref(tab:cost-tbl)). The simple underlying grid is less computationally costly than methods designed to support curvature or switchbacks required in operational road design [@stuckelberger_improved_2007]. The algorithm starts with the landing nearest to an existing road, and updates the existing road network and construction cost surface with each new road segment before moving on to the next nearest landing. The method was developed by Anderson and Nelson [-@anderson_projecting_2004] who noted that the projected network is sensitive to the ordering of targets and recommended a heuristic for connecting the targets and finding optimal branch locations to deal with this sensitivity.

@@

Hardy's method [-@hardy_forest_2019;-@hardy_landis-ii_2023] is similar to our ILCP method in that it breaks the MTAP down into many STAPs and finds the least cost path for each with the ordering of road construction determined by the distance from the existing road network. One difference is that Hardy uses the Dijkstra algorithm to solve a “single-source many-targets shortest path problem”, which is a special case of STAPs that finds the path between the harvested cell and one of multiple potential connection points on the road network. In contrast our algorithm first finds the closest point on the road network to the selected landing point and then uses the Dijkstra algorithm to solve least cost path between the two. However, since the cost of construction on existing roads is zero we expect that in practice these two approaches are similar and the algorithm will select the least cost location for the road to connect to the existing road network. It is possible that a sub-optimal road connection location is chosen by our algorithm when the existing road network is not fully connected and the closest point is on an existing road fragment that is not connected to the optimal part of the road network. Within cutblocks, Hardy's method iterates over each cell in the harvest area and builds a road if it is not within skidding distance of an existing road, while we create landing points to serve as target locations for road construction.

@@

The ILCP and MST methods connect landings within each cutblock to the existing road network, so an important first step is to select one or more landing locations in each block. We investigated the sensitivity of results to the pattern and density of landing locations by comparing results derived from the centroid of each cutblock to regular or randomly selected landing densities of 1 (low) and 10 (high) points per km^2^ (Figure \@ref(fig:proj-meth-fig)). We used default settings in the Hardy QGIS plugin except for skidding distance which was set to 200m. A skidding distance of 200 m corresponds to \~ 9 pts/km^2^. We compared two resolutions of the cost surface by aggregating the 1 ha resolution cost raster to a coarser 100 ha (1000 by 1000 m) resolution. In total, we compared 12 road projection method variants to a cutblocks only scenario with no projected roads (Table \@ref(tab:method-tbl)).

```{r method-tbl}
tbl2 <- "Projection method|Sample type|Sample density (pts/km^2^)|Resolutions
MST|Regular|1|1 ha and 100 ha
MST|Regular|10|1 ha and 100 ha
MST|Random|1|1 ha and 100 ha
MST|Random|10|1 ha and 100 ha
MST|Centroid|N/A|1 ha and 100 ha
ILCP|Regular|10|1 ha 
Hardy QGIS|N/A|~9|1 ha
Cutblocks only|N/A|N/A|1 ha

@@

To help make these road network projection methods more easily and widely useable, we implemented the ILCP and MST methods in the open source R package `roads` [@endicott_roads_2023]. The `roads` package is dependent on several other R packages that made this work possible, including, among others: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @noauthor_igraph_2022], `terra` for manipulating spatial raster data [@hijmans_terra_2023], and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022]. The package is available on CRAN (The Comprehensive R Archive Network). The source code for the package is available on GitHub (<https://github.com/LandSciTech/roads>), along with tutorials and documentation (<https://landscitech.github.io/roads/>).

@@

ilcpProj <- projectAll(tsbs = tsb,
                      paramTable = paramTable %>% 
                        filter(sampleType == "regular") %>% 
                        mutate(method = "ilcp"),
                      costSurface = tsaCost,
                      cutblocks = cutblocks,
                      existingRoads = exRoads,
                      fileLocation = here(data_path_drvd, "..", "for_fig"))

allProj <- bind_rows(mstProj, ilcpProj, .id = "method") %>%
  arrange(desc(sampleType), sampleDens)

@@

allProj <- bind_rows(
  allProj,
  tibble(sampleType = "", sampleDens = 0, method = "Hardy QGIS",
             output = list(here(data_path_drvd, "..", "for_fig", 
                                "HardyQGISTest.gpkg")))) %>% 
  mutate(mapTitle =  paste(ifelse(method == 1, "MST",
                                  ifelse(method == 2, "ILCP", method)),
                           sampleType,
                           ifelse(sampleDens == 1e-06 & 
                                    sampleType != "centroid", "low density", 
                                  ifelse(sampleDens == 1e-05, 
                                         "high density", ""))))

@@

```{r proj-meth-fig, fig.cap="Comparison of minimum spanning tree (MST), iterative least cost paths (ILCP), and the Hardy QGIS plugin road network projection algorithms. Landing locations were selected as the centroid of the cutblock, or a random or regularly spaced sample of points inside each cutblock (Table 2). Target cutblocks are outlined in grey, existing roads are black, projected roads are yellow, observed roads are red. Cost (1 ha resolution) is indicated by grey shading. The large four sided cutblock in the top right was added to more clearly show algorithm behaviour within cutblocks."}
knitr::include_graphics(here("analysis/figures/projection_methods_figure.png"))
```

@@

```{r aspat-perf-fig, fig.height=4, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants within cutblocks, outside cutblocks, and overall (1 ha cost raster). Values greater than zero indicate overprojection and negative values indicate underprojection. Cutblocks not accessible via the observed road network were excluded from analysis. Performance varies among metrics (columns), projection method variants (colours), and within and outside cutblocks."}
# organize data for ggplot
doAspatPlot <- function(tab){
  meanTable_long <- tab %>% dplyr::select(-runTime, -order) %>% 
  pivot_longer(-c(sampleType, sampleDens, areaMean, method),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
                             sampleType),
         sampleDens = paste(method, sampleType, sampleDens) %>% 
           factor(levels = c("mst NA centroid",
                             "NA NA observed",
                             "mst random low sample density",
                             "mst regular low sample density",
                             "mst random high sample density",
                             "mst regular high sample density",
                             "ilcp regular high sample density",
                             "NA NA klementQGIS",
                             "NA NA NA"),
                  labels = c("MST centroid", 
                             "Observed",
                             "MST random low density",
                             "MST regular low density",
                             "MST random high density",
                             "MST regular high density",
                             "ILCP regular high density",
                             "Hardy QGIS",
                             "Cutblocks only")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road\ndensity", "Road\npresence",
                                    "Distance\nto road",
                                    "Forestry\ndisturbance\nfootprint",
                                    "Road\ndisturbance\nfootprint")))

@@

# save runTimes for later comparison
ilcp_time <- meanTableRealCuts %>% 
  filter(method == "ilcp", areaMean == "overall") %>% 
  pull(runTime) %>% round(2)

@@

knitr::include_graphics(here("analysis/figures/Figure4_map_update.png"))

@@

```{r spat-perf-fig, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Aspatial variation of in performance among projection method variants (1 ha cost raster). F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted pixels that were observed and sensitivity is the proportion of observed pixels that were correctly predicted. Values closer to one indicate better performance. Cutblocks not accessible via the observed road network were excluded from analysis."}
doSpatPerf <- function(tab){
  matchData_sum <- tab %>% ungroup() %>%
    mutate(agreement = factor(agreement,
                              levels = c("False negative", "False positive",
                                         "Agree roaded","Agree roadless",
                                         "Pre-existing roads")),
           sampleDens = paste(method, sampleType, sampleDens) %>%
             str_replace("1e-06", "low density") %>%
             str_replace("1e-05", "high density") %>%
             str_replace("centroid low density", "centroid") %>%
             str_replace("klementQGIS NA", "klementQGIS") %>%
             factor(levels = c("mst centroid",
                               "mst random low density",
                               "mst regular low density",
                               "mst random high density",
                               "mst regular high density",
                               "ilcp regular high density",
                               "NA klementQGIS", 
                               "NA cutOnly NA"),
                    labels = c("MST centroid", 
                               "MST random low density",
                               "MST regular low density",
                               "MST random high density",
                               "MST regular high density",
                               "ILCP regular high density",
                               "Hardy QGIS", 
                               "Cutblocks only")),
           metric = factor(metric,
                           levels = c("roadPresence",
                                      "forestryDisturbance",
                                      "roadDisturbance"),
                           labels = c("Road presence",
                                      "Forestry disturbance\nfootprint",
                                      "Road disturbance\nfootprint"))) %>%
    group_by(metric, sampleDens, agreement) %>%
    summarise(count = sum(count), .groups = "drop_last") %>%
    mutate(perc = count/sum(count)*100)

@@

As expected, none of the projection methods yield accurate spatial projections of road presence (i.e. the exact locations of roads) (F-Measure: `r perf_tbl_rng["Road presence"]`). Neither do any of the methods provide better information about the exact location of the forestry disturbance footprint than the locations of cutblocks only (F-Measure: `r perf_tbl_rng["Forestry disturbance\nfootprint"]`). All methods provide comparably good projections of the road disturbance footprint (F-Measure: `r perf_tbl_rng["Road disturbance\nfootprint"]`), though higher sampling densities yield fewer false negatives and more false positives (Figure \@ref(fig:spat-perf-fig)). The ILCP projection method had comparable performance to the MST projection method with the same sampling strategy. The ILCP method was faster to compute however, with a run time of `r ilcp_time` minutes compared to `r mst_time` minutes for the MST method.

<!-- TO DO: Sarah figure out why `r ilcp_time` and `r mst_time` are coming out as NA, and how to fix that.-->

@@

Outside cutblocks, we saw very little variation in performance among projection method variants. All were somewhat better than nothing, but the improvement was small because in this landscape new landings were fairly close to existing roads. Outside cutblocks our ILCP algorithm is similar to Hardy's algorithm, so we expect little difference in behaviour where targets are far from existing roads. Within cutblocks, the density of landing points had a much larger impact on performance than the choice of DCLP or MST algorthm. The Hardy algorithm most accurately projected high road density within blocks, but examination of spatial patterns clarified that the projected tortuousity is quite different from observed switchbacks, and may be an artefact of searching for nearest destinations >200 m away on a 100 m grid. One could increase the density of landing points in the DCLP/MST projections to match the observed high density of roads within cutblocks, but if realistic switchbacks in mountainous terrain is a key goal then none of these algorithms are sufficient.

@@

meanTable <- meanTable %>% mutate(resolution = "100") %>% bind_rows(meanTable2) %>% 
  filter((method != "ilcp"| is.na(method)) & (sampleType != "cutOnly" | is.na(sampleType)))

matchData <- matchData %>% mutate(resolution = "100") %>% bind_rows(matchData2) %>% 
  filter(method != "ilcp", sampleType != "cutOnly")

@@

```{r figA1, fig.height=6, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods for the fine resolutiom (1 ha) and coarse resolution (100 ha) cost rasters. Metrics were summarised within cutblocks, outside cutblocks, and overall. Values greater than 0 indicate the projection overestimated the metric and negative values indicate an underestimation."}
# organize data for ggplot
meanTable_long <- meanTable %>% dplyr::select(-runTime, -order) %>% 
  pivot_longer(-c(sampleType, sampleDens, areaMean, resolution, method),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
                             sampleType),
         sampleDens = paste(method, sampleType, sampleDens) %>% 
           factor(levels = c("mst NA centroid",
                             "NA NA observed",
                             "mst random low sample density",
                             "mst regular low sample density",
                             "mst random high sample density",
                             "mst regular high sample density",
                             "ilcp regular high sample density",
                             "NA klementQGIS", 
                             "NA NA NA"),
                  labels = c("MST centroid", 
                             "Observed",
                             "MST random low density",
                             "MST regular low density",
                             "MST random high density",
                             "MST regular high density",
                             "ILCP regular high density",
                             "Hardy QGIS",
                             "Cutblocks only")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road\ndensity", "Road\npresence",
                                    "Distance\nto road",
                                    "Forestry\ndisturbance\nfootprint",
                                    "Road\ndisturbance\nfootprint")))

@@

matchData_sum <- matchData %>% ungroup() %>%
  mutate(agreement = factor(agreement,
                            levels = c("False negative", "False positive",
                                       "Agree roaded","Agree roadless",
                                       "Pre-existing roads")),
         sampleDens = paste(method, sampleType, sampleDens) %>%
           str_replace("1e-06", "low density") %>%
           str_replace("1e-05", "high density") %>%
           str_replace("centroid low density", "centroid") %>%
           str_replace("klementQGIS NA", "klementQGIS") %>%
           factor(levels = c("mst centroid",
                             "mst random low density",
                             "mst regular low density",
                             "mst random high density",
                             "mst regular high density",
                             "ilcp regular high density",
                             "NA klementQGIS"),
                  labels = c("MST centroid", 
                             "Random low density",
                             "Regular low density",
                             "Random high density",
                             "Regular high density",
                             "ILCP regular high density",
                             "Hardy QGIS")),
         metric = factor(metric,
                         levels = c("roadPresence",
                                    "forestryDisturbance",
                                    "roadDisturbance"),
                         labels = c("Road presence",
                                    "Forestry disturbance\nfootprint",
                                    "Road disturbance\nfootprint"))) %>%
  group_by(metric, sampleDens, agreement, resolution) %>%
  summarise(count = sum(count), .groups = "drop_last") %>%
  mutate(perc = count/sum(count)*100)
